Constraint Grammar as a SAT problem

### 2016-12-23 doc/lrec/eval_futurework.tex
renaming

\paragraph{Full expressivity of CG-3}
As for longer-term goals, we want to handle the full expressivity of CG-3,
with \textsc{map}, \textsc{add},  \textsc{addreading} and
\textsc{substitute} rules, as well as dependency structure. 
This also means finding different kinds of conflicts.
In order to implement rules that may add new readings, or new tags to
existing readings, we need to modify our approach in the SAT-encoding.
Even if the lexicon gives all readings that exist in the lexicon, the
user might give a nonexistent reading, or in the case of MAP, a
syntactic tag, which is (by definition) not in the lexicon. We may need to move
to a more scalable solution.

>>>>

\paragraph{Full expressivity of CG-3}
As for longer-term goals, we want to handle the full expressivity of CG-3,
with \textsc{map}, \textsc{add},  \textsc{append} and
\textsc{substitute} rules, as well as dependency structure. 
This also means finding different kinds of conflicts.
In order to implement rules that may add new readings, or new tags to
existing readings, we need to modify our approach in the SAT-encoding.
Even if the lexicon gives all readings that exist in the lexicon, the
user might give a nonexistent reading, or in the case of {\sc map}, a
syntactic tag, which is (by definition) not in the lexicon. We may need to move
to a more scalable solution.

### 2016-12-23 doc/nodalida/_region_.tex
renaming

%
% File nodalida2015.tex
%
% Contact beata.megyesi@lingfil.uu.se
%
% Based on the instruction file for EACL 2014
% which in turn was based on the instruction files for previous 
% ACL and EACL conferences.

@@

\usepackage{color}
\newcommand{\todo}[1]{{\color{cyan}\textbf{[TODO: }#1\textbf{]}}}
\usepackage{tikz}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) --
  (.25,.15) -- cycle;}
\newcommand{\Mypm}{\mathbin{\tikz [x=1.4ex,y=1.4ex,line width=.1ex] \draw (0.0,0) -- (1.0,0) (0.5,0.08) -- (0.5,0.92) (0.0,0.5) -- (1.0,0.5);}}%

\title{Constraint Grammar as a SAT problem}

% \author{Inari Listenmaa \and Koen Claessen \\
%  Chalmers University of Technology, Gothenburg, Sweden \\
%  {\tt \{inari,koen\}@chalmers.se} }

@@

\begin{document}
\maketitle

@@

\section{Introduction}

We represent Constraint Grammar \cite{karlsson1995constraint} 
as a Boolean satisfiability (SAT) problem.
This is attractive from several reasons: formal logic is
well-studied, and serves as an abstract language to reason about the
properties of CG. Constraint rules encoded in logic capture richer
dependencies between the tags than standard CG.

Applying logic to reductionist grammars has been explored earlier by \cite{lager98,lager_nivre01}, but it was never adopted for use.
% ; logic programming was too slow to be used for tagging or parsing. 
Since those works, SAT solving techniques have improved significantly, and 
they are used in domains such as microprocessor design and computational 
biology---these problems easily match or exceed CG in complexity. 
% Is this true for grammars with thousands of rules?

Due to these advances, we were able to revisit the idea and develop it further. 
% Whereas \cite{lager98} provides a proof-of-concept description, 
Missing from the previous logic-based works, 
we solve eventual rule conflicts by finding a solution that 
discards the least number of rules.
% allows the maximal number of rules to be applied.
We test our implementation by parsing texts in the order of 10,000s--100,000s
words, using grammars with hundreds of rules.

%SAT solving is based on a technique called unit propagation:
%a set of complex Boolean clauses is simplified, starting from unit
%clauses, which consist of just a single variable, and working up to a
%solution, where all variables in the set have a True or False value.

@@

The ambiguous passage can be either a noun phrase, \emph{la}\texttt{<det>} \emph{casa}\texttt{<n>} 
`the house'  or a verb phrase \emph{la}\texttt{<prn>}  \emph{casa}\texttt{<v><pri><p3>} `(he/she) marries her'. 
We add the following rules:

@@

SAT solver performs a search, 
and starts building possible models that satisfy both constraints. 
In addition to the given constraints, we have default rules to emulate
the CG principles: an analysis is true if no rule affects it,
and at least one analysis for each word is true---the notion of
``last'' is not applicable.

With these constraints, we get two solutions. The interaction of the rules regarding \emph{la}  disambiguates the POS of \emph{casa} for free, and ordering of the rules doesn't matter.

\begin{enumerate}
\item [\texttt{1)}]
\begin{verbatim}
"<la>"
        "el" det def f sg
"<casa>"
        "casa" n f sg
\end{verbatim}
\item [\texttt{2)}]
\begin{verbatim}
"<la>"
        "lo" prn p3 f sg
"<casa>"
        "casar" v pri p3 sg
        "casar" v imp p2 sg
\end{verbatim}
\end{enumerate}

%Without more context or additional rules we cannot fully disambiguate the passage,
%but unlike with standard CG, we get only legal combinations in one solution.

@@

\noindent The most important differences between the traditional and the SAT-based approach are summarised below:

@@

\paragraph{Rules are unordered.}

We have experimented with different strategies:
\begin{itemize}
\item Maximise
% \footnote{Something about local maximum}
%The algorithm commits to a subset of clauses (e.g. \emph{all words must have more than 0 true analyses}), and tries to maximise the number of true variables without changing that subset. So it is a local maximum as regards to the given set of clauses. 
number of rules used---We discarded this strategy, because it was too strong: if a rule leads to a conflict in one case, it was not be applied anywhere.
\item Emulate sequential order---Enter clauses produced by each rule one by one, and assume the solver state reached so far is correct; if a conflict is introduced by new clauses, discard them and move on to next.
\item [\checkmark] Maximise number of rule applications---If there is a conflict, find the smallest number of rule applications to discard so that the conflict is solved.
\end{itemize}
The chosen heuristic produced the best results with examples in the order of tens or hundreds of rules.
Larger rule sets are yet to be tested.

@@

\paragraph{Rules disambiguate more.}
Considering our example phrase and rules, standard CG implementation
can only remove readings for the word which is tagged as \texttt{prn} or
\texttt{det}. The SAT-based implementation interprets the rules as
``determiner and verb together are illegal'', and is free to take action that concerns also the word in the condition (\texttt{1 n} or \texttt{1 v}).
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. A rule with more conditions can be written as many
rules, each condition taking its turn to be in position to be removed
or selected. \\

% \begin{itemize}
% \item [] \texttt{REMOVE n IF (-1 prn) ;} \\
%          \texttt{REMOVE v IF (-1 det) ;}
% \end{itemize}

% SAT-based approach gives identical results with both sets of rules,
% whereas the standard CG would remove one reading from \emph{casa} and leave \emph{la} ambiguous.
% Testing this property with more complex rules and larger rule sets remains to be done.

@@

\noindent These three features might change the way that rules are written. 
We predict that less rules are needed; whether this holds in
the order of thousands of rules remains to be tested.
Getting rid of ordering could ease the task of the grammar writer,
since it removes the need to estimate the best sequence of rule
applications.

However, the implementation should not be too different from the
existing ones---our preliminary evaluation against VISL CG-3 shows
promising results in this regard.
\pagebreak

@@

The worst-case complexity of SAT is exponential, whereas the standard
implementations of CG are polynomial, but with advances in SAT solving
techniques, the performance in the average case is much more feasible than in the previous works done in 90s--00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.

Table~\ref{table:time} shows the execution time compared to VISL CG-3. 
The number of rules in the grammar is more significant than the word
count; this holds also for the performance of VISL CG-3.
From the SAT solving side, maximisation is the most costly operation.
Parsing Don Quijote (384,155 words) with the rule set of 261 rules,
the function was called 147,253 times, and with 19 rules, 132,255 times.
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
The difference in the execution times suggests that there are other
reasons for the worse performance---this is to be expected,
SATCG is currently just a naive proof-of-concept implementation with no optimisations.

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|c|}
     \hline
   \textbf{\# words} & \textbf{\# rules} &  \textbf{SATCG} & \textbf{VISL CG-3} \\ \hline
                384,155  & 19   &  13.7s   & 4.2s\\ %\hline
                124,493  & 99   &  55.7s   & 4.8s \\ % \hline
	        384,155  & 130  &  1m14.1s & 6.1s \\ %\hline
                384,155  & 261  &  2m53.7s & 10.7s \\ \hline
  \end{tabular}
  \caption{Execution times.}
  \label{table:time}
\end{table}

\paragraph{Performance against VISL CG-3}

We took a tagged corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged} of 21865 words for Spanish, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We only took select and remove rules (no substitute/iff), after which the constraint grammar had 261 rules.
With this setup, we took the text of the tagged corpus with all
ambiguities, and ran both SATCG and VISL CG-3 with the same grammar. 
Treating the corpus as the gold standard, the disambiguation by
SATCG achieves 75.09 \% correct result and VISL CG-3 75.80 \%.
Similar patterns were observed with small (\textless{}20) rule sets
written by the authors; depending on the subset, SATCG and VISL CG-3 had
a difference of at most $\Mypm$ 1.5 \%. 
Introducing rules one by one up to 19, the
performance improved in a very similar rate, with less than 0.5 \%
difference between the systems at each new rule.

% Additional tests could include plugging SATCG into Apertium
% translation pipeline, and comparing the translation quality.

% \paragraph{Removing rules}
% \begin{table}
%   \centering
%   \begin{tabular}{|l|c|c|}
%      \hline
%     \textbf{\# rules} & \textbf{SATCG} & \textbf{VISL CG-3} \\
%     \hline 
% 	19 &  83.15 & 83.33 \\ \hline  % 19 out of 23 preselected by authors
% 	18 &  83.19 & 83.04 \\ \hline  % 18 out of 19
% 	17 &  83.201 & 83.03 \\ \hline   % 17 out of 18/19, no diff
% 	16 &  83.197 & 83.030 \\ \hline  % 16 out of 17/18, no diff
% 	15 &  83.188 & 83.022\\ \hline  % 15 out of 16/17. Tie with 14 for SATCG, chose to retain the rule that gave better results for VISL CG-3.
% 	14 &  83.180 & 83.005 \\ \hline  % Tie with 15 for SATCG.
% 	13 &  83.159 & 82.951 \\ \hline  % 13 out of 15 
% 	12 &  83.134 & 82.884\\ \hline  % 12 out of 14
% 	11 &  83.089 & 82.847\\ \hline  
% 	10 &  83.051 & 82.802\\ \hline  
% 	9  & 82.997 & 82.802 \\ \hline
% 	8  & 82.906 & 82.681 \\ \hline  % REMOVE:r_imp Imp IF (0 Subj) (-1* CnjSub) ; 
% 	7  & 82.743 & 82.594 \\ \hline  % SELECT:s_pro PrnIndep IF (1C VerbFin) ;
% 	6  & 82.556 & 82.398\\ \hline  %  REMOVE:r_v_v Verb IF (-1C Vblex) ;
% 	5  & 82.327 & 82.028 \\ \hline % SELECT:s_adv Adv IF (1 Cnj_Rel_End OR Prep OR Det)
% 	4  & 82.065 & 82.028 \\ \hline % REMOVE:r_vfin_prep VerbFin IF (0 Prep) (1 Inf) 
% 	3  & 81.387 & 81.233 \\ \hline % REMOVE:r_noun N    IF (-1 N OR Prn)
% 	2  & 80.23 & 79.97 \\ \hline % SELECT:s_pri_1 (pri p3) IF (0 (vblex imp p2 sg))
% 	1  & 76.69 & 76.62 \\ \hline 
% 	0  & 70.70 & 70.70 \\ \hline
%   \end{tabular}
%   \caption{Result of removing rules one by one, and how close to the gold standard they get.}
%   \label{table:results_remove}
% \end{table}
% In this experiment, we took 19 rules written by the authors, which give 83.15 \% correct results with SATCG and 83.33 \% with VISL CG-3.
% Then we removed rules one by one; after getting a score for $n$ rules, we took the combination of $n-1$ rules for which SATCG gives the best score, and we ran the rule set with VISL CG-3 for comparison. Table~\ref{table:results_remove} shows the percentages compared to the gold standard for both systems.

@@

>>>>

\message{ !name(nodalida2015.tex)}%
% File nodalida2015.tex
%
% Contact beata.megyesi@lingfil.uu.se
%
% Based on the instruction file for EACL 2014
% which in turn was based on the instruction files for previous 
% ACL and EACL conferences.

@@

\usepackage{color}
\newcommand{\todo}[1]{{\color{cyan}\textbf{[TODO: }#1\textbf{]}}}
\usepackage{tikz}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
\title{Constraint Grammar as a SAT problem}

\author{Inari Listenmaa \and Koen Claessen \\
  Chalmers University of Technology, Gothenburg, Sweden \\
  {\tt \{inari,koen\}@chalmers.se} }

@@

\begin{document}

\message{ !name(nodalida2015.tex) !offset(-3) }

\maketitle

@@

\section{Introduction}
%Constraint Grammar (CG) was first introduced by \cite{karlsson1995constraint}. 
%It disambiguates output by morphological analyser, 
%and has been used for many tasks in computational linguistics, such as POS tagging,
%surface syntax and machine translation.
%The current state-of-the-art parser VISL CG-3 is implemented (...);
%in addition there are a number of finite-state based implementations \cite{nemeskey14}.

We represent Constraint Grammar \cite{karlsson1995constraint} 
as a Boolean satisfiability (SAT) problem.
Applying logic to reductionist grammars has been explored earlier by 
\cite{lager98,lager_nivre01}, but it was never adopted for use---logic programming was simply 
too slow to be used for tagging or parsing. 
However, modern SAT solvers are fast and used for many applications where constraints must be
satisfied, such as software and hardware verification.
%SAT solving is based on a technique called unit propagation:
%a set of complex Boolean clauses is simplified, starting from unit
%clauses, which consist of just a single variable, and working up to a
%solution, where all variables in the set have a True or False value.

It is attractive to model CG in terms of logic:
these techniques are well-studied, and they capture more dependencies between the tags than the standard CG.

@@

The ambiguous passage can be either a noun phrase, \emph{la\texttt{:det} casa\texttt{:n}} 
`the house'  or a verb phrase \emph{la\texttt{:prn} casa\texttt{:v pri p3}} `(he/she) marries her'. 
We add the following rules:

@@

SAT solver performs a search, 
and starts building possible models that satisfy both constraints. 
In addition to the given constraints, we have default rules to emulate
the CG principles: an analysis is true if no rule affects it,
and at least one analysis for each word is true---the notion of ``last'' is not applicable.
With these constraints, we get two solutions: \texttt{det, n} and \texttt{prn, v pri p3, v imp p2}. 
The interaction of the rules regarding \emph{la}  disambiguates the POS of \emph{casa} 
for free, and ordering of the rules doesn't matter. 
%Without more context or additional rules we cannot fully disambiguate the passage,
%but unlike with standard CG, we get only legal combinations in one solution.

@@

The most important differences between the traditional and the SAT-based approach are summarised below:

@@

\paragraph{Rules are unordered.}
%A typical CG rule sequence starts with rules that are either very common, or have a complex %condition to fulfil.
%Stronger rules appear towards the end, to be applied if none of the weaker rules has. 
%For a rule set such as following, this order is crucial:
%\begin{itemize}
%\item [] \texttt{SELECT Year IF (-1 Seasons OR Months) ;} \\
 %            \texttt{REMOVE Year ;}
%\end{itemize}

We have experimented with different strategies:
\begin{itemize}
\item Maximise
% \footnote{Something about local maximum}
%The algorithm commits to a subset of clauses (e.g. \emph{all words must have more than 0 true analyses}), and tries to maximise the number of true variables without changing that subset. So it is a local maximum as regards to the given set of clauses. 
number of rules used---We discarded this strategy, because it was too strong: if a rule leads to a conflict in one case, it was not be applied anywhere.
\item Emulate sequential order---Enter clauses produced by each rule one by one, and assume the solver state reached so far is correct; if a conflict is introduced by new clauses, discard them and move on to next.
\item [\checkmark] Maximise number of rule applications---If there is a conflict, find the smallest number of rule applications to discard so that the conflict is solved.
\end{itemize}

%combined with the following
%\begin{itemize}
%\item Maximise number of true analyses---Without this heuristic, both solutions to our example %\emph{la casa} are as likely. With this heuristic,  the example would be disambiguated as \texttt{prn v}, because there happens to be two analyses with \texttt{v}, and choosing that, 3 analyses are true, since there is no rule to disambiguate between imperative and indicative.
% We want analyses to be true by default, but still disambiguate as much as possible.
%We have a rule to maximise the number of true analyses; then we take a solution that has the least. 
%\item Minimise number of true analyses---To combat the previous, we want each solution to contain as many true analyses as possible, to ensure that analyses untouched by rules are kept true. %After having multiple solutions which all ensure , we pick the one that disambiguates the most.
%\end{itemize}

The chosen heuristic produced the best results with examples in the order of tens or hundreds of rules.
Larger rule sets (thousands of rules) are yet to be tested.

@@

\paragraph{Rules disambiguate more.}
% Rules interact with each other and disambiguate more than same rules in the standard implementation. 
In standard CG, the rules only tell to remove readings for the word which is tagged as 
\texttt{prn} or \texttt{det}. The SAT-based implementation interprets the rules as ``noun and pronoun together are illegal'', and is free to take action that concerns also the following word.
This means that the rules are logically flipped: compare the rules below to the ones in the example.

\begin{itemize}
\item [] \texttt{REMOVE n IF (-1 prn) ;} \\
             \texttt{REMOVE v IF (-1 det) ;}
\end{itemize}

SAT-based approach gives identical results with both sets of rules,
whereas the standard CG would remove one reading from \emph{casa} and leave \emph{la} ambiguous.
Testing this property with more complex rules and larger rule sets remains to be done.

@@

These three features are likely to change the way that rules are written. \todo{more speculation, examples}

@@

The worst-case complexity of SAT is exponential, whereas the standard implementations 
of CG are polynomial, but with advances in SAT solving techniques, the performance 
in the average case is much more feasible than in the previous works done in 90s--00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.
%We do foreach sentence: foreach rule; 
%whereas VISL-CG3 has foreach rule: foreach sentence.
%the size of the SAT problem is dependent of the number of rules

Table~\ref{table:time} shows the execution time compared to VISL-CG3. 
The number of rules in the grammar is 
Maximisation is the most costly operation
Also, some of the time is explained simply by
SATCG is just a first proof-of-concept implementation and can be optimised.

\begin{table}
  \centering
  \begin{tabular}{|l|l|c|c|}
     \hline
   \textbf{\# words} & \textbf{\# rules} &  \textbf{SATCG} & \textbf{VISL-CG3} \\ \hline
                124,493  & 99    & 55.7s   & 4.8s \\  \hline	                         
                384,155  & 19    &  13.7s  & 4.2s\\ %\hline
                384,155  & 130  &  1m14.1s & 6.1s \\ %\hline
                384,155  & 261  &  2m53.7s & 10.7s \\ \hline
  \end{tabular}
  \caption{Execution times.}
  \label{table:time}
\end{table}

\paragraph{Performance against VISL-CG3}

We took a tagged corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged} of 21865 words for Spanish, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We only took select and remove rules (no substitute/iff), after which the constraint grammar had 261 rules.
With this setup, we took the text of the tagged corpus with all ambiguities, and ran both VISL-CG3 and our implementation with the same grammar. 
Treating the chosen tagset as the gold standard, the disambiguation by SATCG achieves 75.09 \% correct result and VISL-CG3 75.80 \%.

% \paragraph{Removing rules}
% \begin{table}
%   \centering
%   \begin{tabular}{|l|c|c|}
%      \hline
%     \textbf{\# rules} & \textbf{SATCG} & \textbf{VISL-CG3} \\
%     \hline 
% 	19 &  83.15 & 83.33 \\ \hline  % 19 out of 23 preselected by authors
% 	18 &  83.19 & 83.04 \\ \hline  % 18 out of 19
% 	17 &  83.201 & 83.03 \\ \hline   % 17 out of 18/19, no diff
% 	16 &  83.197 & 83.030 \\ \hline  % 16 out of 17/18, no diff
% 	15 &  83.188 & 83.022\\ \hline  % 15 out of 16/17. Tie with 14 for SATCG, chose to retain the rule that gave better results for VISL-CG3.
% 	14 &  83.180 & 83.005 \\ \hline  % Tie with 15 for SATCG.
% 	13 &  83.159 & 82.951 \\ \hline  % 13 out of 15 
% 	12 &  83.134 & 82.884\\ \hline  % 12 out of 14
% 	11 &  83.089 & 82.847\\ \hline  
% 	10 &  83.051 & 82.802\\ \hline  
% 	9  & 82.997 & 82.802 \\ \hline
% 	8  & 82.906 & 82.681 \\ \hline  % REMOVE:r_imp Imp IF (0 Subj) (-1* CnjSub) ; 
% 	7  & 82.743 & 82.594 \\ \hline  % SELECT:s_pro PrnIndep IF (1C VerbFin) ;
% 	6  & 82.556 & 82.398\\ \hline  %  REMOVE:r_v_v Verb IF (-1C Vblex) ;
% 	5  & 82.327 & 82.028 \\ \hline % SELECT:s_adv Adv IF (1 Cnj_Rel_End OR Prep OR Det)
% 	4  & 82.065 & 82.028 \\ \hline % REMOVE:r_vfin_prep VerbFin IF (0 Prep) (1 Inf) 
% 	3  & 81.387 & 81.233 \\ \hline % REMOVE:r_noun N    IF (-1 N OR Prn)
% 	2  & 80.23 & 79.97 \\ \hline % SELECT:s_pri_1 (pri p3) IF (0 (vblex imp p2 sg))
% 	1  & 76.69 & 76.62 \\ \hline 
% 	0  & 70.70 & 70.70 \\ \hline
%   \end{tabular}
%   \caption{Result of removing rules one by one, and how close to the gold standard they get.}
%   \label{table:results_remove}
% \end{table}
% In this experiment, we took 19 rules written by the authors, which give 83.15 \% correct results with SATCG and 83.33 \% with VISL-CG3.
% Then we removed rules one by one; after getting a score for $n$ rules, we took the combination of $n-1$ rules for which SATCG gives the best score, and we ran the rule set with VISL-CG3 for comparison. Table~\ref{table:results_remove} shows the percentages compared to the gold standard for both systems.

@@

\message{ !name(nodalida2015.tex) !offset(-290) }
